A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-25 Manuscript/PNAS-Adams-03-MLC.tex
03-MLC

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
inaccurate.} Computer simulations reveal that for \(\hat{\lambda}\), the
distributional expectations of a Bernoulli variable were mostly upheld.
First, the mean value of \(\hat{\lambda}\) increases as \(\lambda\)
increases, but it is negatively-biased (particularly for small tree
sizes), and is consistently less than the input \(\lambda\) value across
most of its range (Fig. 1 black line). Second, the standard deviation of
\(\hat{\lambda}\) is largest at intermediate values of \(\lambda\) and
smallest at extreme values (Fig.1 red line), implying that the precision
in estimating \(\lambda\) varies across the range of input values.
Additionally, standard deviations of \(\hat{\lambda}\) are negatively
associated with tree size, and for tress of 128 species or less,
\(\hat{\lambda}\) are quite variable, except for cases when \(\lambda\)
is near or equal to \(1\). Third, the distributions of \(\hat{\lambda}\)
are not normal across its range, but become increasingly skewed at more
extreme values of \(\lambda\) (Fig. 1 blue line). For small tree sizes,
it is also clear that distributions are more platykurtic at intermediate
values of \(\hat{\lambda}\). Taken together these results reveal that
\(\hat\lambda\) inconsistently estimates phylogenetic signal, both
across tree sizes and across the range of input values. Additional
simulations (Supporting Information) reveal that incorporating
\(\hat\lambda\) in PGLS ANOVA and regression does not adversely affect
the statistical properties of PGLS parameter estimation or model
evaluation (type I error, power, bias in coefficients). Thus, it is
reasonable to incorporate \(\hat\lambda\) in PGLS as a parameter for
tuning the degree of phylogenetic signal in the dependent variables
during the analysis. However, the statistical properties shown in Fig. 1
demonstrate that \(\lambda\) is unsuitable as an effect size for
measuring the strength of phylogenetic signal in data, and thus
\(\lambda\) should not be used for comparing phylogenetic signal across
datasets.

@@

\textbf{Kappa (\emph{K}) estimates of phylogenetic signal are more
reliable.} Simulation results demonstrate that \emph{K} displays better
statistical properties. First, as expected, mean values of \emph{K}
increase with increasing signal (\(\lambda\)) irrespective of tree size,
albeit non-linearly (Fig. 2 black line). Additionally, the standard
deviation of \emph{K} is consistent across tree sizes (Fig. 2 red line),
and while it increases with \(\lambda\), it is always less than the
mean. This finding is perhaps unsurprising, as \emph{K} is lower-bounded
by 0, and is never large for small values of \(\lambda\). Importantly,
\emph{K} is normally distributed across the range of input \(\lambda\);
a consistent pattern regardless of tree size (Fig. 2 blue line). This
result differs from those of (18), where the skewing appears to be the
result of combining random values generated independently, rather than
being a property of \emph{K} itself. Overall, these findings reveal that
while \emph{K} is more reliable as an estimate of phylogenetic signal,
the non-linear scaling with input signal implies that it should not be
considered an effect size that measures the strength of phylogenetic
signal on a common scale for comparison across datasets.

@@

where \(\theta_{obs}\) is the observed test statistic, \(E(\theta)\) is
its expected value under the null hypothesis, and \(\sigma_\theta\) is
its standard error (32--34). Typically, \(\theta_{obs}\) and
\(\sigma_\theta\) are estimated from the data, while \(E(\theta)\) is
obtained from the distribution of \(\theta\) derived from parametric
theory. However, recent advances in resampling theory (35--38) have
shown that \(E(\theta)\) and \(\sigma_\theta\) may also be obtained from
an empirical sampling distribution of \(\theta\) obtained from
permutation procedures.

@@

where \(K_{obs}\) is the observed phylogenetic signal, and \(\hat\mu_K\)
and \(\hat\sigma_K\) are the mean and standard deviation of the
empirical sampling distribution of \emph{K} obtained via permutation.
The empirical sampling distribution of \emph{K} is first transformed via
a Box-Cox transformation to better adhere to the assumption of
normality.

@@

\begin{figure}
\centering
\includegraphics[width=8.25cm]{{fig.3}.png}
\caption{Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion.{}}
\end{figure}

@@

{[}(A) Linear measures for relative body size, and regions of the body
used to estimate surface area to volume (SA:V) ratios. (B) Permutation
distributions of phylogenetic signal for SA:V and \(\frac{BW}{SVL}\),
with observed values shown as vertical bars. (C) Effect sizes (\(Z_K\))
for SA:V and \(\frac{BW}{SVL}\), with their 95\% confidence intervals
(CI not standardized by \(\sqrt(n)\)).{}{]}\{\{fig.4\}\}.png)

@@

This study is not the first to compare \(\lambda\) and \emph{K} for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \emph{K} values are consistent with those found in
the simulations performed by Münkemüller et al.~(18), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \emph{K} at
intermediate phylogenetic signal strengths was observed (Fig. 2 of ref.
(18)). Recent work by Molina-Venegas and Rodríguez (21) found that
\emph{K} but not \(\lambda\) tended to inflate the estimate of
phylogenetic signal, leading to moderate type I and type II biases, if
polytomic chronograms were used. Their work more thoroughly addressed
previous observations of inflated \emph{K} for incompletely resolved
phylogenetic trees (18, 46). An interesting question is whether an
inflated \emph{K} value leads to an inflated \(Z_K\) or does a tendency
of a particular tree to inflate estimates of \emph{K} also inflate the
values in random permutations of a test, in which case \(Z_K\) is robust
to polytomies? We repeated the analyses in Figs. 1 \& 2, adjusting trees
to have 20\% collapsed nodes \sout{, per the technique of Molina-Venegas
and Rodríguez (21),} and found results were consistent (Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \emph{K} as a descriptive statistic does not inflate
\(Z_K\) as an effect size. Furthermore, because comparison of effect
sizes in a test is a comparison of locations of observed values in their
sampling distributions, which would shift concomitantly because of this
tendency, the \(Z_{12}\) test statistic in equation 4 appears to be
robust in spite of unresolved trees.

>>>>

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
inaccurate.} Computer simulations reveal that for \(\hat{\lambda}\), the
distributional expectations of a Bernoulli variable were mostly upheld.
First, the mean value of \(\hat{\lambda}\) increases as \(\lambda\)
increases. Second, the standard deviation of \(\hat{\lambda}\) is
largest at intermediate values of \(\lambda\) and smallest at extreme
values (Fig.1 red line), implying that the precision in estimating
\(\lambda\) varies across the range of input values. Third, the
distributions of \(\hat{\lambda}\) tend toward normal ddistributions at
intermediate levels of \(\lambda\) but become increasingly skewed at
more extreme values of \(\lambda\) (Fig. 1 blue line). For small tree
sizes, it is also clear that distributions are more platykurtic at
intermediate values of \(\hat{\lambda}\). However, the mean value of
\(\hat{\lambda}\) is negatively-biased (particularly for small tree
sizes but also consistently across most of its range; Fig. 1 black line)
and standard deviations of \(\hat{\lambda}\) are negatively associated
with tree size. For tress of 128 species or less, \(\hat{\lambda}\) are
quite variable, except for cases when \(\lambda\) is near or equal to
\(1\). Taken together these results reveal that \(\hat\lambda\) is a
biased statistic that inconsistently estimates phylogenetic signal, both
across tree sizes and across the range of input values. Additional
simulations (Supporting Information) reveal that incorporating
\(\hat\lambda\) in PGLS ANOVA and regression does not adversely affect
the statistical properties of PGLS parameter estimation or model
evaluation (type I error, power, bias in coefficients). Thus, it is
reasonable to incorporate \(\hat\lambda\) in PGLS as a parameter for
tuning the degree of phylogenetic signal in the dependent variables
during the analysis. However, the statistical properties shown in Fig. 1
demonstrate that \(\lambda\) is unsuitable as an effect size for
measuring the strength of phylogenetic signal in data, and thus
\(\lambda\) should not be used for comparing phylogenetic signal across
datasets.

@@

\textbf{Kappa (\emph{K}) estimates of phylogenetic signal are more
reliable.} Simulation results demonstrate that \emph{K} displays better
statistical properties. First, as expected, mean values of \emph{K}
increase with increasing signal (\(\lambda\)) irrespective of tree size,
albeit non-linearly (Fig. 2 black line). Additionally, the standard
deviation of \emph{K} is consistent across tree sizes (Fig. 2 red line),
and while it increases with \(\lambda\), it is always less than the mean
(low coefficient of variation). This finding is perhaps unsurprising, as
\emph{K} is lower-bounded by 0, and is never large for small values of
\(\lambda\). Importantly, \emph{K} is normally distributed across the
range of input \(\lambda\); a consistent pattern regardless of tree size
(Fig. 2 blue line). This result differs from those of (18), where the
skewing appears to be the result of combining random values generated
independently, rather than being a property of \emph{K} itself. Overall,
these findings reveal that while \emph{K} is more reliable as an
estimate of phylogenetic signal, the non-linear scaling with input
signal implies that it should not be considered an effect size that
measures the strength of phylogenetic signal on a common scale for
comparison across datasets.

@@

where \(\theta_{obs}\) is the observed test statistic, \(E(\theta)\) is
its expected value under the null hypothesis, and \(\sigma_\theta\) is
its standard error (32--34). Typically, \(\theta_{obs}\) and
\(\sigma_\theta\) are estimated from the data, while \(E(\theta)\) is
obtained from the distribution of \(\theta\) derived from parametric
theory. However, recent advances in resampling theory (35--38) have
shown that \(E(\theta)\) and \(\sigma_\theta\) may also be obtained from
an empirical sampling distribution of \(\theta\) simulated from
permutation procedures.

@@

where \(K_{obs}\) is the observed phylogenetic signal, and \(\hat\mu_K\)
and \(\hat\sigma_K\) are the mean and standard deviation of the
empirical sampling distribution of \emph{K} obtained via permutation.
The empirical sampling distribution of \emph{K} can be first transformed
via a Box-Cox transformation to better adhere to the assumption of
normality.

@@

\begin{figure}
\centering
\includegraphics[width=8.25cm]{{fig.3}.png}
\caption{Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion.  Means from simulation runs are shown for comparative ease. Individual values from each simulation run are available in Supporting Informaton. }
\end{figure}

@@

\textbackslash begin\{figure\} \centering
\includegraphics[width=8.25cm]{{fig.4}.png}

\textbackslash caption\{ (A) Linear measures for relative body size, and
regions of the body used to estimate surface area to volume (SA:V)
ratios. (B) Permutation distributions of phylogenetic signal for SA:V
and \(\frac{BW}{SVL}\), with observed values shown as vertical bars. (C)
Effect sizes (\(Z_K\)) for SA:V and \(\frac{BW}{SVL}\), with their 95\%
confidence intervals (CI not standardized by \(\sqrt(n)\)). \}

\textbackslash end\{figure\}

@@

This study is not the first to compare \(\lambda\) and \emph{K} for
their ability as statistics to measure phylogenetic signal. Our results
for \(\lambda\) and \emph{K} values are consistent with those found in
the simulations performed by Münkemüller et al.~(18), but that study
investigated type I error rates and statistical power, finding that
\(\lambda\) performed better in both regards, irrespective of species
number in trees. Although not the central focus of their study, the same
tendency for variable \(\lambda\) and consistent \emph{K} at
intermediate phylogenetic signal strengths was observed (Fig. 2 of ref.
(18)). Recent work by Molina-Venegas and Rodríguez (21) found that
\emph{K} but not \(\lambda\) tended to inflate the estimate of
phylogenetic signal, leading to moderate type I and type II biases, if
polytomic chronograms were used. Their work more thoroughly addressed
previous observations of inflated \emph{K} for incompletely resolved
phylogenetic trees (18, 46). An interesting question is whether an
inflated \emph{K} value leads to an inflated \(Z_K\) or does a tendency
of a particular tree to inflate estimates of \emph{K} also inflate the
values in random permutations of a test, in which case \(Z_K\) is robust
to polytomies? We repeated the analyses in Figs. 1 \& 2, adjusting trees
to have 20\% collapsed nodes, per the technique of Molina-Venegas and
Rodríguez (21), and found results were consistent (Supporting
Information). This confirms that any tendency of incompletely resolved
trees to inflate \emph{K} as a descriptive statistic does not inflate
\(Z_K\) as an effect size. Furthermore, because comparison of effect
sizes in a test is a comparison of locations of observed values in their
sampling distributions, which would shift concomitantly because of this
tendency, the \(Z_{12}\) test statistic in equation 4 appears to be
robust in spite of unresolved trees.
